Judicious Batching: The Impact of Batch Ordering Imaging Tests in the Emergency Department

### 2024-09-09 v2/manuscriptv2.Rmd
getting going

---
title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits...
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
---

@@

@@

To that end, we restrict attention to ED visits for complaints where batching occurs no more than 5 percent of the time (recall that our baseline sample only includes complaints with a $>5$ percent batching rate). We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (6). The results of this exercise are displayed in Appendix Table 1. They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.

@@

\begin{table}[!htbp]
\centering
\caption{Combined 2SLS and Causal Mediation Analysis Results}
\label{tab:combined_results}
\begin{threeparttable}
\begin{tabularx}{\textwidth}{Xcccc}
\toprule
& Number of Imaging Tests & Log LOS & 72-hr Return with Admission \\ \cmidrule(l){2-2} \cmidrule(l){3-3} \cmidrule(l){4-5}
\midrule
\multicolumn{5}{l}{\textit{2SLS Results}} \\
\\
Batched & 0.699** & -0.539 & -0.036 & \\
& (0.286) & (1.534) & (0.034) & \\
\\
  \textit{Baseline Controls} &  $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
  \textit{Testing Inclination} &  $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
  \textit{Imaging Ordered} & $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
\midrule
\multicolumn{5}{l}{\textit{Causal Mediation Analysis}} \\
\\
ACME & & 0.083*  [0.017, 0.15] & -0.003**  [-0.005, 0.00] \\
ADE &  & -0.565  [-3.608, 2.56] & -0.034  [-0.101, 0.03] \\
Total Effect &  & -0.481  [-3.549, 2.60] & -0.037  [-0.104, 0.03] \\
\bottomrule
\end{tabularx}
\end{threeparttable}
\end{table}

@@

\begin{table}[!htbp]
\centering
\caption{Subgroup Analysis Results by Chief Complaint}
\label{tab:subgroup_analysis}
\tiny
\begin{tabularx}{\textwidth}{Xcccccc}
\toprule
Chief Complaint & Batched Coefficient & ACME LOS & ADE LOS & ACME 72hr Return & ADE 72hr Return \\
\midrule
Extremity Complaints & 0.709* & 0.133* & 0.268 & -0.003 & -0.090 \\
 & (0.411) & [-0.030, 0.290] & [-3.640, 3.870] & [-0.012, 0.000] & [-0.214, 0.040] \\
 \\
Falls/MVCs/Trauma & 0.268 & 0.040 & -0.019 & -0.002 & 0.037 \\
 & (0.422) & [-0.103, 0.180] & [-3.109, 2.960] & [-0.009, 0.000] & [-0.055, 0.130] \\
 \\
Dizziness/Syncope & 1.572** & 0.193** & -1.353 & -0.011** & -0.023 \\
 & (0.701) & [0.015, 0.390] & [-4.043, 1.370] & [-0.025, 0.000] & [-0.270, 0.210] \\
 \\
Neurological Issue & 1.652 & 0.139 & -7.353 & 0.003 & -0.220 \\
 & (1.983) & [-0.242, 0.520] & [-16.602, 2.500] & [-0.020, 0.030] & [-0.987, 0.560] \\
 \\
Fatigue and Weakness & 0.793 & 0.072 & -0.286 & -0.002 & 0.010 \\
 & (0.686) & [-0.059, 0.220] & [-1.422, 0.940] & [-0.013, 0.000] & [-0.102, 0.120] \\
 \\
Fevers/Sweats/Chills & -0.181 & -0.025 & -1.086 & 0.002 & 0.061 \\
 & (0.858) & [-0.250, 0.200] & [-3.079, 1.000] & [-0.020, 0.030] & [-0.336, 0.430] \\
\bottomrule
\end{tabularx}
\end{table}

@@

This study provides empirical evidence on the impacts of batch ordering imaging tests in emergency departments, illuminating the trade-offs between comprehensive initial assessments and efficient resource utilization. Our findings, derived from a quasi-experimental design leveraging physician practice variation, align closely with our theoretical model of ED decision-making. We demonstrate that batch ordering, while leading to more thorough initial evaluations, results in increased resource use and longer ED stays.

The fundamental tension we uncover is between the information gain approach of sequential testing and the comprehensive but potentially excessive nature of batch ordering. Our results indicate that batch ordering leads to approximately 70 additional imaging tests per 100 patient encounters compared to a sequential approach. This substantial increase in testing volume translates to longer ED stays but also slightly reduced 72-hour return rates with admission. These findings underscore the complex decision-making environment in EDs, where physicians must balance the need for thorough diagnostics against operational efficiency and patient flow.

Importantly, our subgroup analysis reveals significant heterogeneity in the effects of batch ordering across different presenting complaints. For conditions with high diagnostic uncertainty or potential for serious underlying issues, such as dizziness or syncope, the benefits of comprehensive initial testing may outweigh the costs of increased resource utilization. Conversely, for more straightforward presentations, a sequential, information gain approach appears more appropriate. This variation in effects emphasizes the need for nuanced, complaint-specific approaches to test ordering in emergency settings.

Our results have significant implications for ED management and health policy. They suggest that a one-size-fits-all approach to diagnostic testing in the ED is suboptimal. Instead, ED managers and policymakers should consider implementing targeted batch ordering strategies that account for the specific nature of patient presentations and the relative risks of missed diagnoses versus overcrowding and resource strain. Future research should focus on developing and validating decision support tools that can guide clinicians in choosing between batch and sequential ordering based on patient-specific factors and presenting complaints.

In conclusion, this study contributes to our understanding of ED operations by quantifying the causal impacts of different diagnostic testing strategies. By highlighting the nuanced trade-offs involved in batch ordering, we provide a foundation for more informed decision-making in emergency care. As healthcare systems continue to grapple with the dual challenges of providing high-quality care and managing limited resources, insights from studies like this one will be crucial in shaping evidence-based policies and practices in emergency medicine.

\newpage

>>>>

---
title: "The Impact of Batch Ordering Imaging Tests in the Emergency Department"
authors:
  - name: Jacob Jameson
    department: Interfaculty Initiative in Health Policy
    affiliation: Harvard University
    location: Cambridge, MA
    email: jacobjameson@g.harvard.edu
  - name: Soroush Saghafian
    department: Kennedy School of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: soroush_saghafian@hks.harvard.edu
  - name: Robert Huckman
    department: Harvard Business School
    affiliation: Harvard University
    location: Boston, MA
    email: rhuckman@hbs.edu
  - name: Nicole Hodgson
    department: Department of Emergency Medicine
    affiliation: Mayo Clinic
    location: Phoenix, AZ
    email: Hodgson.Nicole@mayo.edu
abstract: |
  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits...
keywords:
  - Emergency Department
  - Operational Efficiency
  - Diagnostic Testing
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
output: rticles::arxiv_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
---

@@

To determine whether or not the effect of batch tendency on LOS is mediated by the number of tests ordered, we estimate a mediation analysis using the approach outlined by @imai2010general. Our analysis, scrutinizes the pathways through which physician batch tendency influences patient length of stay (LOS) in the emergency department. The analysis delineates both direct effects, which capture the influence of batch tendency on LOS without intermediaries, and indirect effects, which operate through the mediator of test ordering volume. Our findings underscore a significant Average Direct Effect (ADE) of batch tendency on LOS, estimated at -0.15  (p = 0.046).

@@

To that end, we restrict attention to ED visits for complaints where batching occurs no more than 10 percent of the time (recall that our baseline sample only includes complaints with a >2 percent batching rate). We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (3). The results of this exercise are displayed in Appendix Table 1. They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.

@@

\begin{table}[!htbp]
\centering
\caption{Combined 2SLS and Causal Mediation Analysis Results}
\label{tab:combined_results}
\begin{threeparttable}
\begin{tabularx}{\textwidth}{Xcccc}
\toprule
& Number of Imaging Tests & Log LOS & 72-hr Return with Admission \\ \cmidrule(l){2-2} \cmidrule(l){3-3} \cmidrule(l){4-5}
\midrule
\multicolumn{5}{l}{\textit{2SLS Results}} \\
\\
Batched & 0.699** & -0.539 & -0.036 & \\
& (0.286) & (1.534) & (0.034) & \\
\\
  \textit{Baseline Controls} &  $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
  \textit{Testing Inclination} &  $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
  \textit{Imaging Ordered} & $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
\midrule
\multicolumn{5}{l}{\textit{Causal Mediation Analysis}} \\
\\
ACME & & 0.083*  [0.017, 0.15] & -0.003**  [-0.005, 0.00] \\
ADE &  & -0.565  [-3.608, 2.56] & -0.034  [-0.101, 0.03] \\
Total Effect &  & -0.481  [-3.549, 2.60] & -0.037  [-0.104, 0.03] \\
\bottomrule
\end{tabularx}
\begin{tablenotes}
\small
\item Note: ** p < 0.05, * p < 0.1. Full controls included. Estimates from 2SLS and causal mediation analysis, presenting impacts on number of tests, length of stay, and 72-hour return rates.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\begin{table}[!htbp]
\centering
\caption{Subgroup Analysis Results by Chief Complaint}
\label{tab:subgroup_analysis}
\tiny
\begin{tabularx}{\textwidth}{Xcccccc}
\toprule
Chief Complaint & Batched Coefficient & ACME LOS & ADE LOS & ACME 72hr Return & ADE 72hr Return \\
\midrule
Extremity Complaints & 0.709* & 0.133* & 0.268 & -0.003 & -0.090 \\
 & (0.411) & [-0.030, 0.290] & [-3.640, 3.870] & [-0.012, 0.000] & [-0.214, 0.040] \\
 \\
Falls/MVCs/Trauma & 0.268 & 0.040 & -0.019 & -0.002 & 0.037 \\
 & (0.422) & [-0.103, 0.180] & [-3.109, 2.960] & [-0.009, 0.000] & [-0.055, 0.130] \\
 \\
Dizziness/Syncope & 1.572** & 0.193** & -1.353 & -0.011** & -0.023 \\
 & (0.701) & [0.015, 0.390] & [-4.043, 1.370] & [-0.025, 0.000] & [-0.270, 0.210] \\
 \\
Neurological Issue & 1.652 & 0.139 & -7.353 & 0.003 & -0.220 \\
 & (1.983) & [-0.242, 0.520] & [-16.602, 2.500] & [-0.020, 0.030] & [-0.987, 0.560] \\
 \\
Fatigue and Weakness & 0.793 & 0.072 & -0.286 & -0.002 & 0.010 \\
 & (0.686) & [-0.059, 0.220] & [-1.422, 0.940] & [-0.013, 0.000] & [-0.102, 0.120] \\
 \\
Fevers/Sweats/Chills & -0.181 & -0.025 & -1.086 & 0.002 & 0.061 \\
 & (0.858) & [-0.250, 0.200] & [-3.079, 1.000] & [-0.020, 0.030] & [-0.336, 0.430] \\
\bottomrule
\end{tabularx}
\small Note: * p < 0.1, ** p < 0.05. ACME: Average Causal Mediation Effect, ADE: Average Direct Effect. LOS is measured in log-transformed hours. 72hr Return is the probability of return within 72 hours with admission.
\end{table}

@@

### 2024-09-09 v2/manuscriptv2.tex
getting going

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Author template for INFORMS Journal on Data Science (ijds) [interim solution; new styles under construction]
%% Mirko Janc, Ph.D., INFORMS, mirko.janc@informs.org
%% ver. 0.91, March 2015 - updated November 2020 by Matthew Walls, matthew.walls@informs.org
%% Adapted for rticles by Rob J Hyndman Rob.Hyndman@monash.edu. Dec 2021
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[,,nonblindrev]{informs}

\OneAndAHalfSpacedXI
%%\OneAndAHalfSpacedXII % Current default line spacing
%%\DoubleSpacedXII
%%\DoubleSpacedXI

%% BEGIN MY ADDITIONS %%
\usepackage{hyperref}

@@

\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}
\usepackage{caption}

%% END MY ADDITIONS %%

% Natbib setup for author-year style
\usepackage{natbib}
 \bibpunct[, ]{(}{)}{,}{a}{}{,}%
 \def\bibfont{\small}%
 \def\bibsep{\smallskipamount}%
 \def\bibhang{24pt}%
 \def\newblock{\ }%
 \def\BIBand{and}%

%% Setup of theorem styles. Outcomment only one.
%% Preferred default is the first option.
\TheoremsNumberedThrough     % Preferred (Theorem 1, Lemma 1, Theorem 2)
%\TheoremsNumberedByChapter  % (Theorem 1.1, Lema 1.1, Theorem 1.2)
\ECRepeatTheorems

%% Setup of the equation numbering system. Outcomment only one.
%% Preferred default is the first option.
\EquationsNumberedThrough    % Default: (1), (2), ...
%\EquationsNumberedBySection % (1.1), (1.2), ...

% For new submissions, leave this number blank.
% For revisions, input the manuscript number assigned by the on-line
% system along with a suffix ".Rx" where x is the revision number.
\MANUSCRIPTNO{}

%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%

% Outcomment only when entries are known. Otherwise leave as is and
%   default values will be used.
%\setcounter{page}{1}
%\VOLUME{00}%
%\NO{0}%
%\MONTH{Xxxxx}% (month or a similar seasonal id)
%\YEAR{0000}% e.g., 2005
%\FIRSTPAGE{000}%
%\LASTPAGE{000}%
%\SHORTYEAR{00}% shortened year (two-digit)
%\ISSUE{0000} %
%\LONGFIRSTPAGE{0001} %
%\DOI{10.1287/xxxx.0000.0000}%

% Author's names for the running heads
% Sample depending on the number of authors;
\RUNAUTHOR{%
Jameson, Saghafian, Huckman
 and Hodgson
}
% \RUNAUTHOR{Jones and Wilson}
% \RUNAUTHOR{Jones, Miller, and Wilson}
% \RUNAUTHOR{Jones et al.} % for four or more authors
% Enter authors following the given pattern:
%\RUNAUTHOR{}

\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variability in the ED}

\TITLE{To Batch or Not to Batch: Test-Ordering Variability in the ED}

\ARTICLEAUTHORS{%
\AUTHOR{Jacob Jameson}
\AFF{Interfaculty Initiative in Health Policy, Harvard
University, \EMAIL{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}}}

\AUTHOR{Soroush Saghafian}
\AFF{Kennedy School of Government, Harvard
University, \EMAIL{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}}}

\AUTHOR{Robert Huckman}
\AFF{Harvard Business
School, \EMAIL{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}}}

\AUTHOR{Nicole Hodgson}
\AFF{Mayo
Clinic, \EMAIL{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}}}

%
}

\ABSTRACT{We use practice variation across physicians to uncover the
role of test-ordering on care delivery in the ED. Using records of over
45,000 Mayo Clinic emergency department visits\ldots{}}

\KEYWORDS{Emergency Department, Operational Effeciency, Diagnostic
Testing}

\maketitle

@@

Healthcare delivery, particularly in the emergency department (ED),
requires a delicate balance between ensuring optimal patient outcomes
and optimizing resource utilization. Achieving these twin goals
necessitates timely and accurate diagnosis, which enables prompt and
appropriate treatment, improving patient prognosis and reducing the
likelihood of adverse events. Efficient patient discharge from the ED is
crucial for alleviating overcrowding, a severe issue associated with
higher complication rates and increased mortality \citet{bernstein2009}.
In this complex environment, the availability and performance of
diagnostic tests, especially imaging studies, play a pivotal role in
shaping both patient care and operational efficiency \citet{naseim2015}.

A critical question in ED management pertains to whether physicians
should batch order diagnostic imaging tests or order them sequentially.
This decision represents a fundamental tradeoff between potentially
reducing patient length of stay and risking over-testing. Batch
ordering, where multiple imaging tests are ordered simultaneously, may
expedite the diagnostic process but can lead to unnecessary tests,
increased costs, patient anxiety, and potential harm from follow-up of
false-positive results \citet{koch2018}. Conversely, a more sequential
approach might reduce unnecessary testing but could extend a patient's
time in the ED, potentially contributing to overcrowding. Striking the
right balance between comprehensive evaluation and efficient resource
use is crucial, yet research on optimal test ordering strategies in the
ED remains limited \citet{saghafian2015}.

@@

Contrary to most healthcare settings where patients exhibit choice, they
are predominantly passive in their physician assignment in the ED. In
most EDs, however, physicians have discretion in picking their patients.
In contrast, patients arriving at the Mayo Clinic ED are randomly
assigned to physicians via a rotational patient assignment algorithm
\citet{traub2016emergency}, which removes potential selection bias
concerns for our analyses. In essence, barring arrival time and
shift-level variation, the physician-to-patient matching can be deemed
random. Table \ref{tab:wald_test} displays that patient encounters
(regarding chief complaints and emergency severity) are equitably
distributed across physicians within our study's cohort.

@@

To measure physician batch tendency, we use the physician's residualized
leave-out average batch rate. This measure is derived from two steps
following the approaches taken by \citet{doyle2015measuring},
\citet{dobbie2018effects}, and \citet{eichmeyer2022pathways}. First, we
obtain residuals from a regression model, which includes all ED
encounters in our sample period.

@@

Finally, the monotonicity assumption is necessary for interpreting the
coefficient estimates obtained from the IV approach as Local Average
Treatment Effects (LATEs) if there are heterogeneous treatment effects.
It requires that any patient who is (not) batched by a sequencer
(batcher) would also (not) be batched by a batcher (sequencer)
physician. The literature leveraging the judges design typically
performs two informal tests for its implications. The first one provides
that the first stage should be weakly positive for all subsamples
(\citet{dobbie2018effects}). The second implication asserts that the
instrument constructed by leaving out a particular subsample has
predictive power over that same left-out subsample
(\citet{bhuller2020incarceration}). Appendix Table
\ref{tab:monotonicity} presents both of these tests in the two columns
for various subsamples of interest.

@@

@@

To that end, we restrict attention to ED visits for complaints where
batching occurs no more than 5 percent of the time (recall that our
baseline sample only includes complaints with a \(>5\) percent batching
rate). We estimate a reduced-form regression of each main outcome on
physician batch tendency for the subsample, following equation (6). The
results of this exercise are displayed in Appendix Table 1. They show
that in contrast to results for our main sample, the association between
physician tendency to batch and a given outcome is statistically
indistinguishable from zero and much smaller in magnitude for the
samples of patients who visit the ED with health conditions that are
rarely batched.

@@

This study provides empirical evidence on the impacts of batch ordering
imaging tests in emergency departments, illuminating the trade-offs
between comprehensive initial assessments and efficient resource
utilization. Our findings, derived from a quasi-experimental design
leveraging physician practice variation, align closely with our
theoretical model of ED decision-making. We demonstrate that batch
ordering, while leading to more thorough initial evaluations, results in
increased resource use and longer ED stays.

The fundamental tension we uncover is between the information gain
approach of sequential testing and the comprehensive but potentially
excessive nature of batch ordering. Our results indicate that batch
ordering leads to approximately 70 additional imaging tests per 100
patient encounters compared to a sequential approach. This substantial
increase in testing volume translates to longer ED stays but also
slightly reduced 72-hour return rates with admission. These findings
underscore the complex decision-making environment in EDs, where
physicians must balance the need for thorough diagnostics against
operational efficiency and patient flow.

Importantly, our subgroup analysis reveals significant heterogeneity in
the effects of batch ordering across different presenting complaints.
For conditions with high diagnostic uncertainty or potential for serious
underlying issues, such as dizziness or syncope, the benefits of
comprehensive initial testing may outweigh the costs of increased
resource utilization. Conversely, for more straightforward
presentations, a sequential, information gain approach appears more
appropriate. This variation in effects emphasizes the need for nuanced,
complaint-specific approaches to test ordering in emergency settings.

Our results have significant implications for ED management and health
policy. They suggest that a one-size-fits-all approach to diagnostic
testing in the ED is suboptimal. Instead, ED managers and policymakers
should consider implementing targeted batch ordering strategies that
account for the specific nature of patient presentations and the
relative risks of missed diagnoses versus overcrowding and resource
strain. Future research should focus on developing and validating
decision support tools that can guide clinicians in choosing between
batch and sequential ordering based on patient-specific factors and
presenting complaints.

In conclusion, this study contributes to our understanding of ED
operations by quantifying the causal impacts of different diagnostic
testing strategies. By highlighting the nuanced trade-offs involved in
batch ordering, we provide a foundation for more informed
decision-making in emergency care. As healthcare systems continue to
grapple with the dual challenges of providing high-quality care and
managing limited resources, insights from studies like this one will be
crucial in shaping evidence-based policies and practices in emergency
medicine.

\newpage

% Appendix here
% Options are (1) APPENDIX (with or without general title) or
%             (2) APPENDICES (if it has more than one unrelated sections)
% Outcomment the appropriate case if necessary
%
% \begin{APPENDIX}{<Title of the Appendix>}
% \end{APPENDIX}
%
%   or
%
% \begin{APPENDICES}
% \section{<Title of Section A>}
% \section{<Title of Section B>}
% etc
% \end{APPENDICES}

% Acknowledgments here
\ACKNOWLEDGMENT{}

\bibliographystyle{informs2014}
\bibliography{references.bib}

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{Judicious Batching: The Impact of Batch Ordering Imaging Tests in
the Emergency Department}

\author{
    Jacob Jameson
   \\
    Interfaculty Initiative in Health Policy \\
    Harvard University \\
  Cambridge, MA \\
  \texttt{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}} \\
   \And
    Soroush Saghafian
   \\
    Kennedy School of Government \\
    Harvard University \\
  Cambridge, MA \\
  \texttt{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}} \\
   \And
    Robert Huckman
   \\
    Harvard Business School \\
    Harvard University \\
  Boston, MA \\
  \texttt{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}} \\
   \And
    Nicole Hodgson
   \\
    Department of Emergency Medicine \\
    Mayo Clinic \\
  Phoenix, AZ \\
  \texttt{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}} \\
  }

@@

\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}
\usepackage{caption}
\begin{document}
\maketitle

\begin{abstract}
We use practice variation across physicians to uncover the role of
test-ordering on care delivery in the ED. Using records of over 45,000
Mayo Clinic emergency department visits\ldots{}
\end{abstract}

\keywords{
    Emergency Department
   \and
    Operational Efficiency
   \and
    Diagnostic Testing
  }

@@

Healthcare delivery, particularly in the emergency department (ED),
requires a delicate balance between ensuring optimal patient outcomes
and optimizing resource utilization. Achieving these twin goals
necessitates timely and accurate diagnosis, which enables prompt and
appropriate treatment, improving patient prognosis and reducing the
likelihood of adverse events. Efficient patient discharge from the ED is
crucial for alleviating overcrowding, a severe issue associated with
higher complication rates and increased mortality
(\textbf{bernstein2009?}). In this complex environment, the availability
and performance of diagnostic tests, especially imaging studies, play a
pivotal role in shaping both patient care and operational efficiency
(\textbf{naseim2015?}).

A critical question in ED management pertains to whether physicians
should batch order diagnostic imaging tests or order them sequentially.
This decision represents a fundamental tradeoff between potentially
reducing patient length of stay and risking over-testing. Batch
ordering, where multiple imaging tests are ordered simultaneously, may
expedite the diagnostic process but can lead to unnecessary tests,
increased costs, patient anxiety, and potential harm from follow-up of
false-positive results (\textbf{koch2018?}). Conversely, a more
sequential approach might reduce unnecessary testing but could extend a
patient's time in the ED, potentially contributing to overcrowding.
Striking the right balance between comprehensive evaluation and
efficient resource use is crucial, yet research on optimal test ordering
strategies in the ED remains limited (\textbf{saghafian2015?}).

@@

Contrary to most healthcare settings where patients exhibit choice, they
are predominantly passive in their physician assignment in the ED. In
most EDs, however, physicians have discretion in picking their patients.
In contrast, patients arriving at the Mayo Clinic ED are randomly
assigned to physicians via a rotational patient assignment algorithm
(\textbf{traub2016emergency?}), which removes potential selection bias
concerns for our analyses. In essence, barring arrival time and
shift-level variation, the physician-to-patient matching can be deemed
random. Table \ref{tab:wald_test} displays that patient encounters
(regarding chief complaints and emergency severity) are equitably
distributed across physicians within our study's cohort.

@@

To measure physician batch tendency, we use the physician's residualized
leave-out average batch rate. This measure is derived from two steps
following the approaches taken by (\textbf{doyle2015measuring?}),
(\textbf{dobbie2018effects?}), and (\textbf{eichmeyer2022pathways?}).
First, we obtain residuals from a regression model, which includes all
ED encounters in our sample period.

@@

Finally, the monotonicity assumption is necessary for interpreting the
coefficient estimates obtained from the IV approach as Local Average
Treatment Effects (LATEs) if there are heterogeneous treatment effects.
It requires that any patient who is (not) batched by a sequencer
(batcher) would also (not) be batched by a batcher (sequencer)
physician. The literature leveraging the judges design typically
performs two informal tests for its implications. The first one provides
that the first stage should be weakly positive for all subsamples
((\textbf{dobbie2018effects?})). The second implication asserts that the
instrument constructed by leaving out a particular subsample has
predictive power over that same left-out subsample
((\textbf{bhuller2020incarceration?})). Appendix Table
\ref{tab:monotonicity} presents both of these tests in the two columns
for various subsamples of interest.

@@

To determine whether or not the effect of batch tendency on LOS is
mediated by the number of tests ordered, we estimate a mediation
analysis using the approach outlined by (\textbf{imai2010general?}). Our
analysis, scrutinizes the pathways through which physician batch
tendency influences patient length of stay (LOS) in the emergency
department. The analysis delineates both direct effects, which capture
the influence of batch tendency on LOS without intermediaries, and
indirect effects, which operate through the mediator of test ordering
volume. Our findings underscore a significant Average Direct Effect
(ADE) of batch tendency on LOS, estimated at -0.15 (p = 0.046).

@@

To that end, we restrict attention to ED visits for complaints where
batching occurs no more than 10 percent of the time (recall that our
baseline sample only includes complaints with a \textgreater10 percent
batching rate). We estimate a reduced-form regression of each main
outcome on physician batch tendency for the subsample, following
equation (3). The results of this exercise are displayed in Appendix
Table 1. They show that in contrast to results for our main sample, the
association between physician tendency to batch and a given outcome is
statistically indistinguishable from zero and much smaller in magnitude
for the samples of patients who visit the ED with health conditions that
are rarely batched.

@@

\bibliographystyle{unsrt}
\bibliography{references.bib}
